A quick analysis of Baltimore crime
========================================================

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

I'm going to do a very simple analysis of Baltimore crime to show off R. We'll use data downloaded from Baltimore City's awesome open data site (this was downloaded a couple of years ago so if you download now, you will get different results). 

### Getting data

* Arrest data: https://data.baltimorecity.gov/Crime/BPD-Arrests/3i3v-ibrt 
* CCTV data: https://data.baltimorecity.gov/Crime/CCTV-Locations/hdyb-27ak 

Let's load the data:
```{r}
arrest_tab=read.csv("BPD_Arrests.csv", stringsAsFactors=FALSE)
cctv_tab=read.csv("CCTV_Locations.csv", stringsAsFactors=FALSE)

# these columns are mislabeled, so fix them
tmp=arrest_tab$sex
arrest_tab$sex=arrest_tab$race
arrest_tab$race=tmp
```

### Exploring data

```{r}
# dimension of table (data.frame)
dim(arrest_tab)

# what are the columns
names(arrest_tab)

# what is the average arrest age?
mean(arrest_tab$age)

# the range of arrest ages
range(arrest_tab$age)

# how many arrests per sex
table(arrest_tab$sex)

# what are the most common offenses
head(sort(table(arrest_tab$incidentOffense),decreasing=TRUE))

# what are the offenses that only happen once
tab <- table(arrest_tab$incidentOffense)
tab[tab == 1]

# range of arrests after removing those w/ age==0
range(arrest_tab$age[arrest_tab$age>0])
```

Offenses by sex

```{r}
tab <- table(arrest_tab$incidentOffense, arrest_tab$sex)
```

Let's see a table of arrests by sex and race
```{r}
table(sex=arrest_tab$sex,race=arrest_tab$race)
```

A histogram of age

```{r}
hist(arrest_tab$age,nc=100)
with(arrest_tab,hist(age[sex=="M"],nc=100)) # males only
with(arrest_tab,hist(age[sex=="F"],nc=100)) # females only

```

### Are males and females arrested at different ages on average?

Let's take a look at how age depends on sex. Let's plot age as a function of sex first (notice how we indicate that sex is a `factor`). 

```{r}
plot(arrest_tab$age~factor(arrest_tab$sex))
```

One of the neat things about R is that statistical model building and testing is built-in. The model we use is $y_i=\beta_0+\beta_1 x_i$ where $y_i$ is age of sample (example) $i$ and $x_i$ is an indicator variable $x_i \in \{0,1\}$ with $x_i=1$ if the $i$-th record (example) is male. You can check that $\beta_1$ is the difference in mean age between females and males.
We use the formula syntax to build a linear regression model. 

```{r}
# let's ignore those records with missing sex
fit=lm(age~factor(sex),data=arrest_tab,subset=arrest_tab$sex %in% c("M","F"))
summary(fit)
```

We see that $\beta_1 \approx -0.2$ meaning that the arrest age for males is about 2.5 months younger. So there is very little difference in the average age (which is what the linear model is testing) but we see that the probability of observing this difference from a sample of this size **when there is no difference in average age** is small $p \approx 0.01$. Since we have a very large number of examples, or records, this testing framework will declare very small differences as *statistically significant*. We'll return to this theme later in class.


### Geographic distribution of arrests.

First we need to extract latitude and longitude from location, we'll use some string functions to do this

```{r}
tmp=gsub("\\)","",gsub("\\(","",arrest_tab$Location))
tmp=strsplit(tmp,split=",")
arrest_tab$lon=as.numeric(sapply(tmp,function(x) x[2]))
arrest_tab$lat=as.numeric(sapply(tmp,function(x) x[1]))
```

Now let's plot

```{r}
plot(arrest_tab$lon, arrest_tab$lat, xlab="Longitude", ylab="Latitude", main="Arrests in Baltimore")
```

We can also use density estimates to make this nicer:

```{r}
smoothScatter(arrest_tab$lat, arrest_tab$lon, xlab="Latitude", ylab="Longitude", main="Arrests in Baltimore")
```

Let's make this fancier using the `ggplot2` graphics systems and the `maps` package containing map data.

```{r}
library(maps)
library(ggplot2)

balto_map = subset(map_data("county", region="maryland"),subregion=="baltimore city")
plt=ggplot()
plt=plt+geom_polygon(data=balto_map,aes(x=long,y=lat),color="white",fill="gray40")
plt=plt+geom_point(data=arrest_tab,aes(x=lon,y=lat),color="blue",alpha=.1)
print(plt)
```

Now let's add CCTV cameras.

```{r}
tmp=gsub("\\)","",gsub("\\(","",cctv_tab$Location))
tmp=strsplit(tmp,split=",")
cctv_tab$lon=as.numeric(sapply(tmp,function(x) x[2]))
cctv_tab$lat=as.numeric(sapply(tmp,function(x) x[1]))

plt=ggplot()
plt=plt+geom_polygon(data=balto_map,aes(x=long,y=lat),color="white",fill="gray40")
plt=plt+geom_point(data=arrest_tab,aes(x=lon,y=lat),color="blue",alpha=.1)
plt=plt+geom_point(data=cctv_tab,aes(x=lon,y=lat),color="red")
print(plt)
```

### A challenge

Is there any relationship between the number of CCTV cameras and the number of arrests? Divide the city into a grid and plot the number of CCTV cameras vs. the number of arrests.

```{r}
# step 1: divide city intro grid for arrest data
# step 1a: find the range of latitude and longitude
latRange=range(arrest_tab$lat,na.rm=TRUE)
lonRange=range(arrest_tab$lon,na.rm=TRUE)

# step 1b: discretize latitude into 50 bins
latGrid=seq(min(latRange),max(latRange),len=50)
latFactor=cut(arrest_tab$lat,breaks=latGrid)

# now longitude
lonGrid=seq(min(lonRange),max(lonRange),len=50)
lonFactor=cut(arrest_tab$lon,breaks=lonGrid)

# step 1c: make a factor indicating geographic grid location
gridFactor=factor(paste(lonFactor,latFactor,sep=":"))

# step 2: do the same for the cctv data
latFactor=cut(cctv_tab$lat,breaks=latGrid)
lonFactor=cut(cctv_tab$lon,breaks=lonGrid)
cctvGridFactor=factor(paste(lonFactor,latFactor,sep=":"))

arrestTab=table(gridFactor)
cctvTab=table(cctvGridFactor)
m=match(names(cctvTab),names(arrestTab))
plot(arrestTab[m]~factor(cctvTab),xlab="Number of CCTV cameras", ylab="Number of Arrests")
```


### Extra analyses

As part of Project 1 you will add to this analysis. Please use the following template:

#### Mihai Sirbu

What question are you asking?:

I am trying to answer: at what time are most people arrested? 
For this prelimary analysis, I plan on making a plot where
hour is the x-axis and the number of arrest is the y-axis. This 
will produced an "Arrest Timeseries. 

What is the code you use to answer it?:

```{r}

time <- strptime(arrest_tab$arrestTime, "%H:%M")
arrest_tab$hours <- as.numeric(format(time, "%H"))

hours_df <- as.data.frame(table(arrest_tab$hours))
names(hours_df) <- c("hour","count")

g <- ggplot(hours_df, aes(hour, count, group=1))+geom_line(color="blue")+geom_point(color="blue")
g <- g+labs(title = "Arrest Timeseries", x="Time of Day",y="Num of Arrests")
g <- g+scale_x_discrete(breaks=seq(0,23,2))
g <- g + theme(plot.title=element_text(size=16,face="bold"),axis.title.x=element_text(size=16,face="bold"),axis.title.y=element_text(size=16,face="bold"))
g


```

What did you observe?

I had originally thought that there would be very little arrests until 8 pm at which point there would be a giant spike from 8 pm to 5 am. But that was not the case. Instead, the two biggest hours of arrest were 6 pm followed by 10 am (!!). At this point, I'm not entirely sure why that might be. I would be surprised, however, if all offenses followed this exact same pattern. 

#### Aaron Dugatkin

What question are you asking?: I am trying to find out how cameras affect the sorts of crimes in their area, both in reducing certain types of crime, or leading to finding more of other types of crime.

What is the code you use to answer it?:

```{r aarondugatkin}
# modified code from above, to create factors, but remove NA

# added by HCB to restore original arrest table
arrest_tab_original = arrest_tab
#

arrest_tab = arrest_tab[!is.na(arrest_tab$lat) & !is.na(arrest_tab$lon),]
latRange=range(arrest_tab$lat,na.rm=TRUE)
lonRange=range(arrest_tab$lon,na.rm=TRUE)
latGrid=seq(min(latRange),max(latRange),len=50)
latFactor=cut(arrest_tab$lat,breaks=latGrid)
lonGrid=seq(min(lonRange),max(lonRange),len=50)
lonFactor=cut(arrest_tab$lon,breaks=lonGrid)
gridFactor=factor(paste(lonFactor,latFactor,sep=":"))
latFactor=cut(cctv_tab$lat,breaks=latGrid)
lonFactor=cut(cctv_tab$lon,breaks=lonGrid)
cctvGridFactor=factor(paste(lonFactor,latFactor,sep=":"))
arrestTab=table(gridFactor)
cctvTab=table(cctvGridFactor)
#count crimes in areas with and without camera
arrestOnCamera = gridFactor %in% names(cctvTab)
count_crime_tab <- table(arrest_tab$incidentOffense, arrestOnCamera)
#merge the two tables, and calculate the difference in crime frequency in the two situations
crime_tab <- data.frame(count_crime_tab[,1], count_crime_tab[,2])
colnames(crime_tab)[1] <- "noCamCrimes"
colnames(crime_tab)[2] <- "camCrimes"
crime_tab$names <- rownames(crime_tab)
crime_tab$campct <- crime_tab$camCrimes/sum(crime_tab$camCrimes)*100
crime_tab$nocampct <- crime_tab$noCamCrimes/sum(crime_tab$noCamCrimes)*100
crime_tab$pctchange <- crime_tab$campct - crime_tab$nocampct
#display the change in crime frequency with crime name in descending order, with the most increased (caught) crimes first
crime_tab <- crime_tab[with(crime_tab, order(-pctchange)), ]
options(scipen=999)
subset(crime_tab, select=c("pctchange"))

# added by HCB to restore original arrest table
arrest_tab = arrest_tab_original
```

What did you observe? The results were interesting. We see a large increase in charges of  narcotics, which may be due to camera surveillance. We also see a decrease in assault, which may be due to the perpetrators of such crimes realizing the dangers of committing such crimes in front of a camera. However, the vast majority of crimes do not even see a 1% change between the two situations, so it would appear as though, overall, cameras do not have a major affect on criminal activity.

#### Krishna Pai

What question are you asking?: Do police officers go out of their way to arrest more black people than white people?

What is the code you use to answer it?:

```{r }
library(ggplot2)
arrest_tab=read.csv("BPD_Arrests.csv", stringsAsFactors=FALSE)
tmp=arrest_tab$sex
arrest_tab$sex=arrest_tab$race
arrest_tab$race=tmp
police_tab=read.csv("Police_Stations.csv", stringsAsFactors=FALSE)
tmp=gsub("\\)","",gsub("\\(","",arrest_tab$Location))
tmp=strsplit(tmp,split=",")
arrest_tab$lon=as.numeric(sapply(tmp,function(x) x[2]))
arrest_tab$lat=as.numeric(sapply(tmp,function(x) x[1]))
tmp=gsub("\\)","",gsub("\\(","",police_tab$Location))
tmp=strsplit(tmp,split=",")
police_tab$lon=as.numeric(sapply(tmp,function(x) x[2])) 
police_tab$lat=as.numeric(sapply(tmp,function(x) x[1])) 
plt=ggplot()
plt=plt+geom_point(data=arrest_tab[arrest_tab$race=='B',],aes(x=lon,y=lat),color="black",alpha=.1)
plt=plt+geom_point(data=arrest_tab[arrest_tab$race=='W',],aes(x=lon,y=lat),color="white",alpha=.1)
plt=plt+geom_point(data=police_tab,aes(x=lon,y=lat),color="red")
print(plt)
```

<<<<<<< HEAD
What did you observe? I was surprised to find that most of the arrests made some distance away from the cluster of police stations were for white people. It would be interesting to investigate what kinds of crimes might have been committed so far away from the stations, and why white people stand out as being arrested at that distance, especially in the east.
=======
What did you observe? 
  2011 had the largest number of arrests but there was no dramatic difference between the number of arrests for each year.  There were 52,868 arrests in 2011 which was 1208 more than the number of arrests for 2012.  


#### Rentao Wu

What are you asking:

I wanted to know if the ratio of female to male crime rates are similar accross the difference races.

What is the code you used to answer it?

```{r}
mytab = table(race=arrest_tab$race, sex=arrest_tab$sex)
#mydf$V1 = NULL
mydf = as.data.frame.matrix(mytab) 
mydf$ratio = mydf$F/mydf$M
mydf$ratio <- round(mydf$ratio, 3)
mydf = mydf[-1,]
mydf
mydf$race <- c("A","B","H","I","U","W")

ggplot(data=mydf, aes(x=race, y=ratio, fill=race)) + geom_bar(stat="identity")
```

What are your observations?
I found out that for most race, the ratio of female to male crime rates is about 0.2. This tells us that there are about 1 female for ever 5 male criminal offenses. I also saw that the female to male crime ratio for the white population is about 0.43 which is much higher than the others. 

#### Krishna Pai

What question are you asking?: Do police officers go out of their way to arrest more black people than white people?

What is the code you use to answer it?:

```{r }
library(ggplot2)
arrest_tab=read.csv("BPD_Arrests.csv", stringsAsFactors=FALSE)
tmp=arrest_tab$sex
arrest_tab$sex=arrest_tab$race
arrest_tab$race=tmp
police_tab=read.csv("Police_Stations.csv", stringsAsFactors=FALSE)
tmp=gsub("\\)","",gsub("\\(","",arrest_tab$Location))
tmp=strsplit(tmp,split=",")
arrest_tab$lon=as.numeric(sapply(tmp,function(x) x[2]))
arrest_tab$lat=as.numeric(sapply(tmp,function(x) x[1]))
tmp=gsub("\\)","",gsub("\\(","",police_tab$Location))
tmp=strsplit(tmp,split=",")
police_tab$lon=as.numeric(sapply(tmp,function(x) x[2])) 
police_tab$lat=as.numeric(sapply(tmp,function(x) x[1])) 
plt=ggplot()
plt=plt+geom_point(data=arrest_tab[arrest_tab$race=='B',],aes(x=lon,y=lat),color="black",alpha=.1)
plt=plt+geom_point(data=arrest_tab[arrest_tab$race=='W',],aes(x=lon,y=lat),color="white",alpha=.1)
plt=plt+geom_point(data=police_tab,aes(x=lon,y=lat),color="red")
print(plt)
```

What did you observe? I was surprised to find that most of the arrests made some distance away from the cluster of police stations were for white people. It would be interesting to investigate what kinds of crimes might have been committed so far away from the stations, and why white people stand out as being arrested at that distance, especially in the east.
>>>>>>> upstream/master
